Mendelian randomization for causal inference in Genome-wide association studies

### 2022-01-28 mr-review.Rmd
more text...

Mendelian randomization with one instrument uses a single genetic marker 
as the proxy ("instrument" or "instrumental variable") for the "exposure" and asks 
whether variation in the genotypes at the genetic marker causes variation in the 
"outcome" variable, which is typically a second complex trait. When the genetic marker 
is a single nucleotide polymorphism on an autosome, the cohort is partitioned into three 
classes, two homozygote classes and a single heterozygote class, based on genotypes 
for that marker. When the three assumptions of mendelian randomization are satisfied, 
estimation of the causal effect of the exposure on the outcome is possible.

@@

Three assumptions are central to Mendelian Randomization. Satisfying them is necessary to infer causal effects in a MR study. Careful evaluation of the assumptions is needed. Briefly, the first assumption states that the exposure is associated with the outcome. The second requires independence, conditional on the exposure and all measured and unmeasured confounders, between the genetic marker and the outcome. The last of the three assumptions states that the genetic marker is independent of all (measured and unmeasured) confounders of the relationship between exposure and outcome. We next discuss these in more detail and consider their evaluation in MR studies.

The first assumption, about the association between the genetic marker and the exposure, holds, but this association is weak, then an amplification of biases, such as those due to violations of assumptions 2 and 3, may result [@vanderweele_methodological_2014].

@@

When the three MR assumptions hold, one may use instrumental variable methods, 
with the genetic marker data, the exposure data, and the outcome data, to estimate the 
effect of the exposure on the outcome.

>>>>

We begin our discussion of Mendelian randomization by considering the simplest case, where there is a single genetic marker and a single outcome variable. Mendelian randomization with one instrument uses a single genetic marker 
as the proxy ("instrument" or "instrumental variable") for the "exposure" and asks 
whether variation in the genotypes at the genetic marker causes variation in the 
"outcome" variable, which is typically a second complex trait. When the genetic marker 
is a single nucleotide polymorphism on an autosome, the cohort is partitioned into three 
classes, two homozygote classes and a single heterozygote class, based on genotypes 
for that marker. When the three assumptions of mendelian randomization are satisfied, 
estimation of the causal effect of the exposure on the outcome is possible.

The outcome variable can be binary, a count, or a continuous variable.

@@

Three assumptions are central to Mendelian Randomization. Satisfying them is necessary to infer causal effects in a MR study. Careful evaluation of the assumptions is needed in MR studies. Briefly, the first assumption states that the exposure is associated with the outcome. The second requires independence, conditional on the exposure and all measured and unmeasured confounders, between the genetic marker and the outcome. The last of the three assumptions states that the genetic marker is independent of all (measured and unmeasured) confounders of the relationship between exposure and outcome. We next discuss these in more detail and consider their evaluation in MR studies.

If the first assumption, about the association between the genetic marker and the exposure, holds, but this association is weak, then an amplification of biases, such as those due to violations of assumptions 2 and 3, may result [@vanderweele_methodological_2014].

@@

The third assumption, sometimes called the ____, ensures that ___.

When the three MR assumptions hold, one may use instrumental variable statistical methods, 
with the genetic marker data, the exposure data, and the outcome data, to estimate and to test the causal effect of the exposure on the outcome.

## Assumption violations

There are several ways that real data may violate the three MR assumptions. These include ___, ____, and ___.

Recognition that the assumptions may be violated has inspired methods advances that 
enable assumption relaxations under certain conditions.

## Statistical models for MR with one instrument and one outcome

## Statistical methods for MR with one instrument and one outcome

The approaches to MR can be classified into two categories: two-stage methods and 
likelihood-based methods. Both two-stage strategies and likelihood-based strategies 
make parametric assumptions for the identification of causal effects [@burgess_review_2017].

Two-stage methods use two regressions. In the first stage, the exposure is regressed on
the instrument to obtain fitted values. In the second stage, the outcome is regressed on the fitted values for the 
exposure, which result from the first regression. Note that the uncertainty in the fitted values of 
the first regression is not considered when performing the second regression.
